---
title: "DATA 621 Blog 2: Timing Operations in R"
author: "Ilya Kats"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

R has many ways of accomplishing the same thing - base functions, various packages. Sometimes the data set is small and there is plenty of time. Other times, there is a lot of data and computational time makes a difference. Below are some ways of timing operations in R, so that you can compare various approaches to find the most optimal. '

## Data Set

First let us generate a random data set. Since this is just about timing any data set will do. Below creates 1,000,000 random observations - large enough to take a bit of time if you manipulate it.

```{r}
x <- runif(1000000, 0, 100)
y <- rnorm(1000000, 0, 10)
```

## `tictoc` Package

```{r}
library(tictoc)
```

```{r}
tic("Linear regression")
l1 <- lm(y ~ x)
toc(log = TRUE)
```

```{r}
tic("Linear regression - outer")
tic("Linear regression - inner")
l1 <- lm(y ~ x)
toc()
l2 <- lm(y ~ x)
toc()
```


## References

Alexej's Blog, 5 ways to measure running time of R code, https://www.r-bloggers.com/5-ways-to-measure-running-time-of-r-code/
